\LaTeX\ Author Guidelines for ICCV Proceedings

### 2021-03-16 chapters/supplementary.tex
Rephrase proofs in appendix

\paragraph{Implementation} Our implementation of \algname{} is based on an union-find data structure and a heap allowing deletion of its elements. 
In Phases 2 and 3, \algname{} is equivalent to a standard hierarchical agglomerative clustering algorithm with complexity $\mathcal{O}(N^2 \log N)$. In Algorithm \ref{detailed_alg}, we show our implementation of Phase 1, involving cannot-link constraints.
In phase 1, the algorithm starts with each node assigned to its own cluster and sorts all edges $e\in E$ in a heap/priority queue (PQ) by their absolute weight $|\cost_e|=|w_e^+ - w_e^-|$ in descending order, so that the most attractive and the most repulsive interactions are processed first. It then iteratively pops one edge $e_{uv}$ from PQ and, depending on the priority $\tilde{\cost}_{uv}$, does the following: in case of attractive interaction $\tilde{\cost}_{uv}>0$, provided that $e_{uv}$ was not flagged as a cannot-link constraint, merge the connected clusters, perform an edge contraction of $e_{uv}$ in $\tilde{\mathcal{G}}_\Pi$ and update the priorities of new double edges as explained in Fig. \ref{fig:edge_contraction_and_contr_graph}. 
If, on the other hand, the interaction is repulsive ($\tilde{\cost}_{uv}\leq 0$) and the option \texttt{addCannotLinkContraints} of Alg. \ref{detailed_alg} is \texttt{True}, then the edge $e_{uv}$ is flagged as cannot-link constraint.

@@

\subsection{Proofs of Propositions \ref{prop:absmax_mutex}, \ref{prop:weight_shift_invariant}, \ref{prop:ultraMetric1}, and \ref{prop:ultraMetric2}  }
\label{sec:proposition_proofs}
% \subsubsection{Mutex Watershed and AbsMax linkage}\label{sec:appendix_abs_max}
\begin{lemma} \label{lemma:absMax_and_complete_property}
Let us consider the \algname{} Algorithm \ref{main_alg} with AbsMax or Complete linkage criteria (see linkage definition in Table \ref{tab:linkage-criteria}). If the algorithm enforces a constraint between two clusters in \emph{Phase 1}, then the interaction between the clusters will never become positive over the course of the following agglomeration steps. Moreover, we also have that with any of these two linkages: i) \algname{} returns the same final clustering whether or not cannot-link constraints are enforced; ii) no clusters are merged in \emph{Phase 2} (after constraints have been removed). 
\end{lemma}
\begin{proof}
First, we note that two clusters are constrained in \emph{Phase 1} only if their interaction is repulsive. We then also observe that with complete linkage (see Table~\ref{tab:linkage_criteria_explicit}) the signed interaction between two clusters can only decrease over the course of the agglomeration. Thus, if two clusters are constrained by the algorithm, their negative interaction cannot increase and become positive later on in the agglomeration process.

If, on the other hand, we consider \algname{} with a Absolute Maximum linkage (see Table~\ref{tab:linkage_criteria_explicit}), during the agglomeration the interaction between two clusters can only increase in absolute value. Thus, the negative interaction $\interact{}(S_i,S_j)<0$ between two constrained clusters can possibly become positive over the course of next agglomeration steps only if there is at least another pair of clusters in the graph that has a positive interaction $\interact{}(S_l,S_t)>0$ higher in absolute value: $|\interact{}(S_l,S_t)| > |\interact{}(S_i,S_j)|$.
If such clusters $S_l, S_t$ with positive interaction exist, we note that they must also be constrained (in the opposite case, the algorithm would have already merged them before to constrain $S_i$ and $S_j$, because their priority is higher). In other words, a constrained negative interaction can become positive only if there is already another positive constrained interaction: but this can never be the case because initially all constrained interactions are negative.

Finally, we prove that \algname{} with either AbsMax or Complete linkage returns the same final clustering whether or not cannot-link constraints are enforced. In Algorithm \ref{detailed_alg}, two clusters are merged only if the condition at line 9 is satisfied (i.e. when an interaction is both positive and not constrained). However, we have just shown that with these two linkage methods an interaction can never be both positive and constrained at the same time, so we directly conclude that the constrained and unconstrained versions of the algorithm will perform precisely the same agglomeration steps.
\end{proof}

@@

The same is not true for \algname{} with Sum linkage criteria (GAEC and HCC-Sum), because by adding a constant $\alpha$ to all edge weights $w_e$, the interaction between two clusters $S_i$ and $S_j$ is increased by a factor $\alpha |E_{ij}|$, which depends on the number of edges  $|E_{ij}|$ connecting the two clusters. Thus, when all edge weighs of the graph are shifted, the agglomeration order may change, as we show in the counter-example of \TODO{FIG}.

The constrained versions of \algname{} (HCC-Avg and HCC-Single) are also not weight-shift invariant: here, the algorithm merges or constrains clusters in a given order, depending on the absolute interactions $|\interact{}(S_i, S_j)|$ between clusters; so, when edge weights are shifted by a constant $\alpha$, the sorting by absolute value can change arbitrarily together with the agglomeration order, as we show in the counter-example of \TODO{FIG}.
Similarly, the Mutex Watershed algorithm is not weight-shift invariant because it uses a linkage criterion that compares weights by their absolute values (see again counter-example in \TODO{FIG}).
\end{proof}

@@

\begin{restatable}{prop}{firstUltraMetricProperty}
\label{prop:ultraMetric1}
Consider a graph $\mathcal{G}(V,E,w_e)$, a linkage criterion $\interact{}$, and an agglomerative algorithm returning a binary rooted tree $T$ with height $h_T$. Then, $(V, d_{T})$ defined in Eq.~\ref{eq:UM_def} is an ultrametric if and only if the following is true:
 % iff the interaction $\interact{}_{T}(u, v)$ associated to the least common ancestor ${(u\vee v)\in T}$ is both lower than the interaction of any other $(u\vee v)$-descendant node and higher than the interaction of any other $(u\vee v)$-ancestor node in the tree $T$. In formulae, this is equivalent to saying that, $\forall u,v,t \in V$:
\begin{equation}\label{eq:UM_assumption}
 h_T(u \vee v) < h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) \geq \interact{}_{T}(u,t) \quad \forall u,v,t \in V.
 % h_T(u \vee v) < h_T(u \vee t) \Leftrightarrow \interact{}_{T}(u,v) > \interact{}_{T}(u,t). 
\end{equation}
In words, condition \ref{eq:UM_assumption} means: if the algorithm merges nodes $u,v$ before to merge nodes $u,t$, then the signed interaction $\interact{}_{T}(u,v)$ between $u$ and $v$ has to be higher or equal than $\interact{}_{T}(u,t)$.
% In words, $(V, d_{T})$ is an ultrametric iff the interaction $\interact{}_{T}(u, v)$ associated to the least common ancestor ${(u\vee v)\in T}$ is both lower than any other $(u\vee v)$-descendant node and higher than any other $(u\vee v)$-ancestor node in the tree $T$.
\end{restatable}
\begin{proof}
% Let us first analyze assumption \ref{eq:UM_assumption}. From $(\Rightarrow)$, we have that:
% \begin{equation}\label{eq:assumption_side_1}
% h_T(u \vee v) < h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) > \interact{}_{T}(u,t). 
% \end{equation}
% In words, if the algorithm first merges node $v$ together with node $u$ and later on merges node $t$ with node $u$, then the signed interaction $\interact{}_{T}(u,v)$ between $u$ and $v$ has to be stronger than $\interact{}_{T}(u,t)$. On the other hand, the $(\Leftarrow)$ side of Eq.~\ref{eq:UM_assumption} can be rewritten as:
% \begin{equation}\label{eq:assumption_side_2}
% h_T(u \vee v) \geq h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) \leq \interact{}_{T}(u,t).
% \end{equation}
% % which in words means: if node $t$ was merged with $u$ at the same or at an earlier iteration  together with node $u$ and later on merges node $t$ with node $u$, then the interaction $\interact{}_{T}(u,v)$

From the definition of $d_{T}$, we have that:
\begin{align}
d_{T}(u,u) &= 0 \qquad &\forall u\in V \label{eq:dist_1} \\ 
d_{T}(u,v) &\geq 0 \qquad &\forall u,v \in V \label{eq:dist_2}\\
d_{T}(u,v)& =d_{T}(v,u) \qquad &\forall u,v \in V \label{eq:dist_3}
\end{align}
 % $d_{T}(u,v)=d_{T}(v,u)$, $d_{T}(u,u)= 0$, and $d_{T}(u,v)\geq 0$ for each $u,v \in V$.
Thus, to show that $(V, d_{T})$ is an ultrametric, we only need to prove the ultrametric property:
\begin{equation}\label{eq:UM_property_original}
d_T(u,v) \leq \max \{d_T(u,t), d_T(v,t)\} \quad \forall u,v,t \in V.
\end{equation}
When at least two of the three nodes $u,v,t \in V$ are the same, this property follows from Eq.~\ref{eq:dist_1} and Eq.~\ref{eq:dist_2}. When nodes $u,v,t\in V$ are distinct, from the definition of $d_{T}$ it follows that Eq.~\ref{eq:UM_property_original} is equivalent to:
\begin{equation}\label{eq:UM_property}
\interact{}_{T}(u,v) \geq \min \{\interact{}_{T}(u,t), \interact{}_{T}(v,t)\}.
\end{equation}
In the following, we prove both sides of the \emph{if and only if} statement in the proposition. First, we prove the $(\Leftarrow)$ side, i.e. that if assumption \ref{eq:UM_assumption} holds, then $(V, d_{T})$ is an ultrametric and \ref{eq:UM_property} holds.

Case 1: in Eq.~\ref{eq:UM_property}, $t\in V$ is part of the sub-tree ${T[u \vee v]}$. In other words, the algorithm first merges node $t$ with either node $u$ or $v$, and then $u$ and $v$ are merged together. Let us assume that $t$ is first merged with $u$ (the following proof also holds for the opposite case in which $t$ is first merged with $v$):
\begin{equation}
h_T(u \vee t) < h_T(u \vee v) = h_T(v \vee t).
\end{equation}
Thus, from assumption (\ref{eq:UM_assumption}) we have that
\begin{equation}
\interact{}_{T}(u \vee t) \geq \interact{}_{T}(v \vee t) \quad  \text{and} \quad \interact{}_{T}(u \vee v) = \interact{}_{T}(v \vee t)
\end{equation}
and Eq.~\ref{eq:UM_property} follows (becoming an equality in this case).

@@

Next, we are left to prove the $(\Rightarrow)$ side of the \emph{if and only if} statement: if $(V, d_{T})$ is an ultrametric, then assumption \ref{eq:UM_assumption} holds.
To prove this statement, we first rephrase it in the following equivalent form: if assumption \ref{eq:UM_assumption} does not hold, then $(V, d_{T})$ is not an ultrametric and \ref{eq:UM_property} does not hold. The negation of assumption \ref{eq:UM_assumption} is
\begin{equation}
h_T(u \vee v) < h_T(u \vee t) \quad \text{and} \quad  \interact{}_{T}(u,v) < \interact{}_{T}(u,t).
\end{equation}
The first condition, in words, is again assuming that the algorithm first merges nodes $u$ and $v$, and later it also merges node $t$ with the cluster containing $u$ and $v$. Thus, we can rephrase this assumption as:
\begin{equation}
\interact{}_{T}(u \vee v) < \interact{}_{T}(u \vee t) = \interact{}_{T}(v \vee t).
\end{equation}
From this, it follows that
\begin{equation}
\interact{}_{T}(u \vee v) < \min \{\interact{}_{T}(u \vee t), \interact{}_{T}(v \vee t)\},
\end{equation}
which is exactly the negation of the ultrametric property \ref{eq:UM_property}.
\end{proof}

@@

Finally, the remaining not constrained versions of \algname{} (HC-Avg, HC-Single, and HC-Complete) satisfy both conditions, so they define a ultrametric, confirming the well-known results of related work in hierarchical clustering on unsigned graphs \cite{johnson1967hierarchical,milligan1979ultrametric}.

@@

\subsection{Mutex Watershed on SSBM graphs}
\begin{prop}
Consider a graph generated by an Erd\H os-R\'enyi signed stochastic block model (SSBM) as described in Section \ref{sec:clustering_problems}, with $N$ nodes, edges added with probability $p$, sign-flip probability $\eta$, $k$ ground-truth clusters, and edge weights Gaussian-distributed with standard deviation $\sigma$. Then, at every iteration, \algname{} with Absolute Maximum linkage (or, in other words, the Mutex Watershed algorithm) always makes a mistake with at least probability $\eta$. 
\end{prop}
\begin{proof}
Thanks to Lemma \ref{lemma:absMax_and_complete_property} we know that \algname{} with Absolute Maximum linkage returns the same clustering whether or not cannot-link-constraints are used. Thus, in the following, we prove the proposition considering the version enforcing constraints.
Let us consider a generic iteration of the algorithm, where two clusters $S_\alpha$ and $S_\beta$ have the highest priority and are popped from priority queue. Then, the MWS algorithm will either merge or constrain them depending on the fact that their interaction $\interact_{\mathrm{AbsMax}}(S_\alpha,S_\beta)$ is  positive or negative (note that, with AbsMax linkage, an interaction can never be positive and constrained, as shown in Lemma \ref{lemma:absMax_and_complete_property}).
By construction of the SSBM, every edge $e\in E$ in the graph has a absolute weight distributed as $|w_e|\sim \mathcal{N}(1,\sigma^2)$. Thus, every edge $e'\in(S_{\alpha} \times S_{\beta})\cap E$ connecting the two clusters has the same probability to have the highest absolute weight, and the sign of the interaction $\interact_{\mathrm{AbsMax}}(S_\alpha,S_\beta)$ will only depend on the sign of this highest edge. Therefore, the probability that the MWS merges two clusters is simply given by the fraction of positive weighted edges connecting them.

Let $\tilde{\Pi}=\{\tilde{S}_1,\ldots,\tilde{S}_k\}$ denote the ground truth clustering, and $\tilde{S}_{\alpha i}=S_\alpha\cap \tilde{S}_i$ denote the intersection between cluster $S_{\alpha}$ and a ground-truth cluster $\tilde{S}_i$.
If the generated graph is dense and $p=1$, then the total number of edges connecting clusters $S_{\alpha}$ and $S_{\beta}$ that have a true attractive or repulsive weight is (according to the ground truth labels):
\begin{equation}
\NBE^+=\sum_{i=1}^k |\tilde{S}_{\alpha i}||\tilde{S}_{\beta i}|, \quad \NBE^-=\sum_{i=1}^k \sum_{j=1,j\neq i}^k |\tilde{S}_{\alpha i}||\tilde{S}_{\beta j}|.
\end{equation}
When we edges in the graph are randomly added with a probability $p$, then the actual number of true attractive and repulsive interactions connecting the two clusters is (according to the ground truth labels):
\begin{equation}
\nBE^+ \sim \mathcal{B}(\NBE^+,p), \qquad \nBE^- \sim \mathcal{B}(\NBE^-,p), 
\end{equation}
where $\mathcal{B}(\NBE,p)$ is the binomial distribution:
\begin{equation}
\mathcal{B}(\nBE;\NBE,p) = \frac{\NBE !}{\nBE! (\NBE-\nBE)!} p^\nBE (1-p)^{\NBE-\nBE}.
\end{equation}
Here, we only assume that $\nBE^+ + \nBE^- >0$, i.e. there is at least one edge connecting the two clusters (otherwise their interaction would be zero and the MWS would not have popped them from priority queue).

So far we have been talking about attractive and repulsive connections according to the ground truth labels. The actual number of attractive interactions connecting the two clusters will be instead given by the sum of the true attractive interactions $\nBE^+_{\mathrm{nf}}\sim \mathcal{B}(\nBE^+,1-\eta)$ that have not been flipped, plus the true negative interactions $\nBE^-_{\mathrm{f}}\sim \mathcal{B}(\nBE^-,\eta)$ that have been flipped.
Putting everything together, given two clusters with $\nBE^+$ true attractive interactions and $\nBE^-$ true negative ones, the highest-absolute-weight edge connecting them has the following probability to be positive:
\begin{align}
 \mathbb{P}[\interact_{\mathrm{AbsMax}}&(S_\alpha,S_\beta)>0;\nBE^+,\nBE^-] = \nonumber \\
=&\sum_{\nBE^+_{\mathrm{nf}}=0}^{\nBE^+}\sum_{\nBE^-_{\mathrm{f}}=0}^{\nBE^-} 
\mathcal{B}(\nBE^-_{\mathrm{f}};\nBE^-,\eta) \mathcal{B}(\nBE^+_{\mathrm{nf}};\nBE^-,1-\eta)\cdot  \nonumber\\
& \qquad \qquad  \cdot \left(\frac{\nBE^+_{\mathrm{nf}}+\nBE^-_{\mathrm{f}}}{\nBE^+ + \nBE^-} \right) \nonumber\\
\stackrel{(*)}{=}&\frac{\nBE^+(1-\eta)+\nBE^-\eta}{\nBE^+ + \nBE^-}
\end{align}  
where in $(*)$ we used the fact that the expected value of a binomial distribution $\mathcal{B}(\nBE,\eta)$ is $\nBE \eta$.

>>>>

\paragraph{Implementation} Our implementation of \algname{} is based on an union-find data structure and a heap allowing deletion of its elements. 
In Phases 2 and 3, \algname{} is equivalent to a standard hierarchical agglomerative clustering algorithm with complexity $\mathcal{O}(N^2 \log N)$. In Algorithm \ref{detailed_alg}, we show our implementation of phase 1, involving cannot-link constraints.
In phase 1, the algorithm starts with each node assigned to its own cluster and sorts all edges $e\in E$ in a heap/priority queue (PQ) by their absolute weight $|\cost_e|=|w_e^+ - w_e^-|$ in descending order, so that the most attractive and the most repulsive interactions are processed first. It then iteratively pops one edge $e_{uv}$ from PQ and, depending on the priority $\tilde{\cost}_{uv}$, does the following: in case of attractive interaction $\tilde{\cost}_{uv}>0$, provided that $e_{uv}$ was not flagged as a cannot-link constraint, merge the connected clusters, perform an edge contraction of $e_{uv}$ in $\tilde{\mathcal{G}}_\Pi$ and update the priorities of new double edges as explained in Fig. \ref{fig:edge_contraction_and_contr_graph}. 
If, on the other hand, the interaction is repulsive ($\tilde{\cost}_{uv}\leq 0$) and the option \texttt{addCannotLinkContraints} of Alg. \ref{detailed_alg} is \texttt{True}, then the edge $e_{uv}$ is flagged as cannot-link constraint.

@@

\subsection{Proofs of Propositions \ref{prop:absmax_mutex}, \ref{prop:weight_shift_invariant}, \ref{prop:ultraMetric1}, and \ref{prop:ultraMetric2}}
\label{sec:proposition_proofs}

\begin{lemma} \label{lemma:absMax_and_complete_not_positive}
If \algname{} Algorithm \ref{main_alg} with \textbf{Complete linkage criteria} enforces a constraint between two clusters in \emph{Phase 1}, then the interaction between the clusters will never become positive over the course of the following agglomeration steps.
\end{lemma}
\begin{proof}
Two clusters are constrained in \emph{Phase 1} only if their interaction is repulsive and, with complete linkage,  the signed interaction between two clusters can only decrease over the course of the agglomeration. Thus, if two clusters are constrained by the algorithm, their negative interaction cannot increase and become positive later on in the agglomeration process.
\end{proof}

\begin{lemma} \label{lemma:absMax_and_complete_not_positive}
If \algname{} Algorithm \ref{main_alg} with \textbf{AbsMax linkage criteria} enforces a constraint between two clusters in \emph{Phase 1}, then the interaction between the clusters will never become positive over the course of the following agglomeration steps.
\end{lemma}
\begin{proof}
During the agglomeration the interaction between two clusters can only increase in absolute value. Thus, the negative interaction $\interact{}(S_i,S_j)<0$ between two constrained clusters can possibly become positive over the course of next agglomeration steps only if there is at least another pair of clusters in the graph that has a positive interaction $\interact{}(S_l,S_t)>0$ higher in absolute value: $|\interact{}(S_l,S_t)| > |\interact{}(S_i,S_j)|$.
If such clusters $S_l, S_t$ with positive interaction exist, we note that they must also be constrained (in the opposite case, the algorithm would have already merged them before to constrain $S_i$ and $S_j$, because their priority is higher). In other words, a constrained negative interaction can become positive only if there is already another positive constrained interaction: but this can never be the case because initially all constrained interactions are negative.
\end{proof}

\begin{lemma} \label{lemma:absMax_and_complete_property}
In the \algname{} Algorithm \ref{main_alg} with AbsMax or Complete linkage criteria (see linkage definition in Table \ref{tab:linkage-criteria}), the same final clustering is returned whether or not cannot-link constraints are enforced.
\end{lemma}
\begin{proof}
In phase 1 of Algorithm \ref{detailed_alg}, two clusters are merged only if the condition at line 9 is satisfied (i.e. when an interaction is both positive and not constrained). From Lemma \ref{lemma:absMax_and_complete_not_positive} and Lemma \ref{lemma:absMax_and_complete_not_positive} follows that with Complete and AbsMax linkage an interaction can never be both positive and constrained at the same time, so we directly conclude that the constrained and unconstrained versions of the algorithm will perform precisely the same agglomeration steps in phase 1.
In phase 2 (after constraints have been removed) no clusters are merged because all interactions are already negative (whether they previously constrained or not). Thus, both constrained and unconstrained versions of \algname{} return the same clustering $\Pi^*$.
\end{proof}

@@

The same is not true for \algname{} with Sum linkage criteria (GAEC and HCC-Sum), because by adding a constant $\alpha$ to all edge weights $w_e$, the interaction between two clusters $S_i$ and $S_j$ is increased by a factor $\alpha |E_{ij}|$, which depends on the number of edges  $|E_{ij}|$ connecting the two clusters. Thus, when all edge weighs of the graph are shifted, the agglomeration order may change. For a simple example of this, it is enough to consider the toy graph in Fig.~\ref{fig:intro_figure}a and shift the weights of the graph by $\alpha=-3$ (see Fig.~\ref{fig:counter_examples_shift_inv_GAEC}).

The constrained versions of \algname{} (HCC-Avg and HCC-Single) are also not weight-shift invariant: here, the algorithm merges or constrains clusters in a given order, depending on the absolute interactions $|\interact{}(S_i, S_j)|$ between clusters; so, when edge weights are shifted by a constant $\alpha$, the sorting by absolute value can change arbitrarily together with the agglomeration order, as we show in the counter-example of Fig.~\ref{fig:counter_examples_shift_inv}.
Similarly, the Mutex Watershed algorithm is not weight-shift invariant because it uses a linkage criterion that compares weights by their absolute values (see again counter-example in Fig.~\ref{fig:counter_examples_shift_inv}.
\end{proof}

@@

\begin{restatable}{prop}{firstUltraMetricProperty}
\label{prop:ultraMetric1}
Consider a graph $\mathcal{G}(V,E,w_e)$, a linkage criterion $\interact{}$, and an agglomerative algorithm returning a binary rooted tree $T$ with height $h_T$. Then, $(V, d_{T})$ defined in Eq.~\ref{eq:UM_def} is an ultrametric if and only if the following is true:
 % iff the interaction $\interact{}_{T}(u, v)$ associated to the least common ancestor ${(u\vee v)\in T}$ is both lower than the interaction of any other $(u\vee v)$-descendant node and higher than the interaction of any other $(u\vee v)$-ancestor node in the tree $T$. In formulae, this is equivalent to saying that, $\forall u,v,t \in V$:
\begin{gather}
\forall u,v,t \in V \nonumber \\
 h_T(u \vee v) < h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) \geq \interact{}_{T}(u,t) \label{eq:UM_assumption}
 % h_T(u \vee v) < h_T(u \vee t) \Leftrightarrow \interact{}_{T}(u,v) > \interact{}_{T}(u,t). 
\end{gather}
In words, condition \ref{eq:UM_assumption} means: if the algorithm merges nodes $u,v$ before to merge nodes $u,t$, then the signed interaction $\interact{}_{T}(u,v)$ between $u$ and $v$ has to be higher or equal than $\interact{}_{T}(u,t)$.
% In words, $(V, d_{T})$ is an ultrametric iff the interaction $\interact{}_{T}(u, v)$ associated to the least common ancestor ${(u\vee v)\in T}$ is both lower than any other $(u\vee v)$-descendant node and higher than any other $(u\vee v)$-ancestor node in the tree $T$.
\end{restatable}
\begin{proof}
% Let us first analyze assumption \ref{eq:UM_assumption}. From $(\Rightarrow)$, we have that:
% \begin{equation}\label{eq:assumption_side_1}
% h_T(u \vee v) < h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) > \interact{}_{T}(u,t). 
% \end{equation}
% In words, if the algorithm first merges node $v$ together with node $u$ and later on merges node $t$ with node $u$, then the signed interaction $\interact{}_{T}(u,v)$ between $u$ and $v$ has to be stronger than $\interact{}_{T}(u,t)$. On the other hand, the $(\Leftarrow)$ side of Eq.~\ref{eq:UM_assumption} can be rewritten as:
% \begin{equation}\label{eq:assumption_side_2}
% h_T(u \vee v) \geq h_T(u \vee t) \Rightarrow \interact{}_{T}(u,v) \leq \interact{}_{T}(u,t).
% \end{equation}
% % which in words means: if node $t$ was merged with $u$ at the same or at an earlier iteration  together with node $u$ and later on merges node $t$ with node $u$, then the interaction $\interact{}_{T}(u,v)$

From the definition of $d_{T}$, it follows that:
\begin{align}
d_{T}(u,u) &= 0 \qquad &\forall u\in V \label{eq:dist_1} \\ 
d_{T}(u,v) &\geq 0 \qquad &\forall u,v \in V \label{eq:dist_2}\\
d_{T}(u,v)& =d_{T}(v,u) \qquad &\forall u,v \in V. \label{eq:dist_3}
\end{align}
 % $d_{T}(u,v)=d_{T}(v,u)$, $d_{T}(u,u)= 0$, and $d_{T}(u,v)\geq 0$ for each $u,v \in V$.
In order to show that $(V, d_{T})$ is an ultrametric, we only need to prove the ultrametric property:
\begin{equation}\label{eq:UM_property_original}
d_T(u,v) \leq \max \{d_T(u,t), d_T(v,t)\} \quad \forall u,v,t \in V.
\end{equation}
When at least two of the three nodes $u,v,t \in V$ are the same, this property follows from Eq.~\ref{eq:dist_1} and Eq.~\ref{eq:dist_2}. When nodes $u,v,t\in V$ are distinct, from the definition of $d_{T}$ it follows that Eq.~\ref{eq:UM_property_original} is equivalent to:
\begin{equation}\label{eq:UM_property}
\interact{}_{T}(u,v) \geq \min \{\interact{}_{T}(u,t), \interact{}_{T}(v,t)\}.
\end{equation}
In the following, we prove both sides of the \emph{if and only if} statement in the proposition. First, we prove the $(\Leftarrow)$ side, i.e. that if assumption \ref{eq:UM_assumption} holds, then $(V, d_{T})$ is an ultrametric and \ref{eq:UM_property} holds.

Case 1: in Eq.~\ref{eq:UM_property}, $t\in V$ is part of the sub-tree ${T[u \vee v]}$. In other words, the algorithm first merges node $t$ with either node $u$ or $v$, and then $u$ and $v$ are merged together. Let us assume that $t$ is first merged with $u$ (the following proof also holds for the opposite case in which $t$ is first merged with $v$):
\begin{equation}\label{eq:case_1}
h_T(u \vee t) < h_T(u \vee v) = h_T(v \vee t).
\end{equation}
Thus, by combining the last equation with assumption (\ref{eq:UM_assumption}), it follows that
\begin{equation}
\interact{}_{T}(u \vee t) \geq \interact{}_{T}(v \vee t) \quad  \text{and} \quad \interact{}_{T}(u \vee v) = \interact{}_{T}(v \vee t)
\end{equation}
and Eq.~\ref{eq:UM_property} follows (becoming an equality in this case).

@@

Next, we are left to prove the $(\Rightarrow)$ side of the \emph{if and only if} statement: if $(V, d_{T})$ is an ultrametric, then assumption \ref{eq:UM_assumption} holds.
To prove this statement, we first rephrase it in the following equivalent form: if assumption \ref{eq:UM_assumption} does not hold, then $(V, d_{T})$ is not an ultrametric and \ref{eq:UM_property} does not hold. If we negate assumption \ref{eq:UM_assumption}, there must be at least three $u,v,t \in V$ such that: 
\begin{equation}
h_T(u \vee v) < h_T(u \vee t) \quad \text{and} \quad  \interact{}_{T}(u,v) < \interact{}_{T}(u,t).
\end{equation}
The first condition, in words, is again assuming that the algorithm first merges nodes $u$ and $v$, and later it also merges node $t$ with the cluster containing $u$ and $v$. Thus, we can rephrase this assumption as:
\begin{equation}
\interact{}_{T}(u \vee v) < \interact{}_{T}(u \vee t) = \interact{}_{T}(v \vee t).
\end{equation}
From this, it follows that
\begin{equation}
\interact{}_{T}(u \vee v) < \min \{\interact{}_{T}(u \vee t), \interact{}_{T}(v \vee t)\},
\end{equation}
which is exactly the negation of the ultrametric property \ref{eq:UM_property}.
\end{proof}

@@

Finally, the remaining not constrained versions of \algname{} (HC-Avg, HC-Single, and HC-Complete) satisfy both conditions, so they define an ultrametric, confirming the well-known results of related work in hierarchical clustering on unsigned graphs \cite{johnson1967hierarchical,milligan1979ultrametric}.

@@

\begin{figure}[t]
% \begin{subfigure}[t]{0.48\textwidth}
\centering
\includegraphics[width=0.48\textwidth,trim=0in 0in 0in 0in,clip]{./figs/counter-examples/counter-example-GAEC.pdf}
% \caption{Original graph and dendrogram}
% \end{subfigure}

% \vspace{2em}
% \begin{subfigure}[t]{0.48\textwidth}
% \centering
% \includegraphics[width=\textwidth,trim=0in 0in 0in 0in,clip]{./figs/counter-examples/counter-example-GAEC-shifted.pdf}
% \caption{Graph with shifted edge-weights ($\alpha=-3$) and GAEC dendrogram}
% \end{subfigure}
        \caption{ Counter-example showing that GAEC is not weight-shift invariant.
        } \label{fig:counter_examples_shift_inv_GAEC}
\end{figure}
\begin{figure}[t!]
% \begin{subfigure}[t]{0.48\textwidth}
\centering
\includegraphics[width=0.48\textwidth,trim=0in 0in 0in 0in,clip]{./figs/counter-examples/counter-example.pdf}
% \caption{Original graph and dendrogram}
% \end{subfigure}

% \vspace{2em}
% \begin{subfigure}[t]{0.48\textwidth}
% \centering
% \includegraphics[width=\textwidth,trim=0in 0in 0in 0in,clip]{./figs/counter-examples/counter-example-shifted.pdf}
% \caption{Graph with shifted edge-weights ($\alpha=+2$) and dendrogram}
% \end{subfigure}
        \caption{ Counter-example showing that HCC-Sum, MWS, HCC-Avg, and HCC-Single are not weight-shift invariant.
        } \label{fig:counter_examples_shift_inv}
\end{figure}

\subsection{Mutex Watershed on SSBM graphs}
\begin{prop} \label{prop:MWS_on_SSBM}
Consider a graph generated by an Erd\H os-R\'enyi signed stochastic block model (SSBM) as described in Section \ref{sec:clustering_problems}, with $N$ nodes, edges added with probability $p$, sign-flip probability $\eta<0.5$, $k$ ground-truth clusters, and edge weights Gaussian-distributed with standard deviation $\sigma$. Then, at every iteration, \algname{} with Absolute Maximum linkage (or, in other words, the Mutex Watershed algorithm) always makes a mistake with at least probability $\eta$. 
\end{prop}
\begin{proof}
Thanks to Lemma \ref{lemma:absMax_and_complete_property} we know that \algname{} with Absolute Maximum linkage returns the same clustering whether or not cannot-link-constraints are used. Thus, in the following, we prove the proposition considering the version enforcing constraints.
Let us consider a generic iteration of the algorithm, where two clusters $S_\alpha$ and $S_\beta$ have the highest priority and are popped from priority queue. Then, the MWS algorithm will either merge or constrain them depending on the fact that their interaction $\interact_{\mathrm{AbsMax}}(S_\alpha,S_\beta)$ is  positive or negative (note that, with AbsMax linkage, an interaction can never be positive and constrained, as shown in Lemma \ref{lemma:absMax_and_complete_property}).
By construction of the SSBM, every edge $e\in E$ in the graph has a absolute weight distributed as $|w_e|\sim \mathcal{N}(1,\sigma^2)$. Thus, every edge $e'\in(S_{\alpha} \times S_{\beta})\cap E$ connecting the two clusters has the same probability to have the highest absolute weight, and the sign of the interaction $\interact_{\mathrm{AbsMax}}(S_\alpha,S_\beta)$ will only depend on the sign of this highest edge. Therefore, the probability that the MWS merges two clusters is simply given by the fraction of positive weighted edges connecting them.

Let $\tilde{\Pi}=\{\tilde{S}_1,\ldots,\tilde{S}_k\}$ denote the ground truth clustering, and $\tilde{S}_{\alpha i}=S_\alpha\cap \tilde{S}_i$ denote the intersection between cluster $S_{\alpha}$ and a ground-truth cluster $\tilde{S}_i$.
If the generated graph is dense, i.e. $p=1$, then the total number of edges connecting clusters $S_{\alpha}$ and $S_{\beta}$ that have a true attractive or repulsive weight is (according to the ground truth labels)
\begin{equation}
\NBE^+=\sum_{i=1}^k |\tilde{S}_{\alpha i}||\tilde{S}_{\beta i}|, \quad \NBE^-=\sum_{i=1}^k \sum_{j=1,j\neq i}^k |\tilde{S}_{\alpha i}||\tilde{S}_{\beta j}|.
\end{equation}
When the edges in the graph are randomly added with a probability $p$, then the actual number of true attractive and repulsive interactions connecting the two clusters is (according to the ground truth labels):
\begin{equation}
\nBE^+ \sim \mathcal{B}(\NBE^+,p), \qquad \nBE^- \sim \mathcal{B}(\NBE^-,p), 
\end{equation}
where $\mathcal{B}(\NBE,p)$ is the binomial distribution:
\begin{equation}
\mathcal{B}(\nBE;\NBE,p) = \frac{\NBE !}{\nBE! (\NBE-\nBE)!} p^\nBE (1-p)^{\NBE-\nBE}.
\end{equation}
Here, we only assume that $\nBE^+ + \nBE^- >0$, i.e. there is at least one edge connecting the two clusters (otherwise their interaction would be zero and the MWS would not have popped them from priority queue).
